---
title: "Differential expression report"
output:
  html_document:
    toc: true
    toc_depth: 2
    mathjax: null
params:
  rmd: ""
---

```{r setup, include = FALSE}
# Significance results and DE figures (scripts/de_significance.R) + model summary -> one DE report
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
source("scripts/common.R")

criterion <- snakemake@params$criterion
figures   <- snakemake@input$figures
counts    <- read_tsv(snakemake@input$summary, show_col_types = FALSE)
model     <- read_tsv(snakemake@input$model, show_col_types = FALSE)

significant <- list.files(snakemake@input$significant, "\\.tsv$", full.names = TRUE) |>
  map(\(f) read_tsv(f, col_select = protein_id, col_types = cols(.default = "c"))) |>
  list_rbind()
n_significant <- n_distinct(significant$protein_id)
n_top <- min(N_TOP_HITS, n_significant)

overview <- tribble(
  ~item, ~value,
  "Dataset", snakemake@wildcards$dataset,
  "Significant if", describe_criterion(criterion),
  "Significant in any contrast", str_glue("{n_significant} proteins")
) |>
  bind_rows(model)
```

## Summary

```{r}
knitr::kable(overview, col.names = NULL)
```

```{r}
knitr::kable(select(counts, label, contrast, up, down, total), col.names = c("contrast", "file name", "up", "down", "total"))
```

`r figure_image(figures, "counts")`

## Volcano plots {.tabset .tabset-dropdown}
Pick a contrast. Dashed lines mark the significance cutoffs; the most significant proteins are labelled.

```{r, results = "asis"}
for (i in seq_len(nrow(counts))) {
  cat("\n\n### ", counts$label[i], "\n\n", figure_image(figures, "volcano", counts$contrast[i]), "\n", sep = "")
}
```

## Top hits
Intensities of the `r n_top` most significant proteins, z-scored per protein (red = above the protein's mean across samples). Samples are grouped by condition.
`r if (n_top < 2) "Not shown: fewer than two significant proteins."`

`r if (n_top >= 2) figure_image(figures, "top_hits")`

## p-value distributions
A healthy contrast shows a flat histogram with a peak near 0 (the real differences). A hump in the middle or near 1 points to a problem with the model, such as an unaccounted batch or pairing.

`r figure_image(figures, "pvalues")`

## How to read this

- **A vs B**: positive log2FC means higher in A. A is the condition listed first in `samples.tsv`; to flip a comparison, reorder the rows of `Input/<dataset>/samples.tsv`. Names shown here come from its `label` column; files are named by condition (A_vs_B).
- **Significant**: `r describe_criterion(criterion)`. padj is the Benjamini-Hochberg adjusted p-value within each contrast.
- **Statistics**: one limma model for all conditions, so every contrast borrows variance information from all samples.
- **Paired samples**: when `samples.tsv` has a `block` column, samples with the same value (e.g. the same mouse) are treated as related, which removes subject-to-subject differences from the comparisons.
- **Tables**: all proteins with statistics in `de/all_proteins/`, significant ones in `de/<criterion>/significant/`.
- **Figures**: each plot is also in its own folder with its data and a script to restyle it: `de/<criterion>/figures/<plot>/`, volcano plots in `figures/volcano/<A_vs_B>/`.
