---
title: "Differential expression report"
output:
  html_document:
    toc: true
    toc_depth: 2
    mathjax: null
params:
  rmd: ""
---

```{r setup, include = FALSE}
# limma statistics + one significance criterion (+ intensities for the heatmap) -> one DE report
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 9)
source("scripts/common.R")

criterion <- snakemake@params$criterion
samples   <- read_tsv(snakemake@input$samples, show_col_types = FALSE)
proteins  <- read_tsv(snakemake@input$intensities, show_col_types = FALSE)
model     <- read_tsv(snakemake@input$model, show_col_types = FALSE)
pairs     <- condition_pairs(samples)
de        <- read_de(snakemake@input$stats, samples) |> map(\(x) call_significance(x, criterion))
counts    <- count_significant(de)

significant <- de |>
  map(\(x) filter(x, direction != "ns")) |>
  list_rbind()

overview <- tribble(
  ~item, ~value,
  "Dataset", snakemake@wildcards$dataset,
  "Significant if", describe_criterion(criterion),
  "Significant in any contrast", str_glue("{n_distinct(significant$protein_id)} proteins")
) |>
  bind_rows(model)

# Top hits: the 50 significant proteins with the smallest p-value in any contrast
top_ids <- significant |>
  summarise(pvalue = min(pvalue), .by = protein_id) |>
  slice_min(pvalue, n = 50, with_ties = FALSE) |>
  pull(protein_id)

conditions <- fct_inorder(samples$condition)
condition_colours <- set_names(scales::hue_pal()(nlevels(conditions)), levels(conditions))

contrast_height <- max(2.5, 0.3 * length(de))
pvalue_height   <- max(2.5, 1.8 * ceiling(length(de) / 4))
heatmap_height  <- max(4, 0.14 * length(top_ids) + 2)
```

## Summary

```{r}
knitr::kable(overview, col.names = NULL)
```

```{r}
knitr::kable(counts)
```

```{r, fig.height = contrast_height}
counts |>
  pivot_longer(c(up, down), names_to = "direction", values_to = "n") |>
  mutate(n = if_else(direction == "down", -n, n), contrast = fct_rev(fct_inorder(contrast))) |>
  ggplot(aes(n, contrast, fill = direction)) +
  geom_col() +
  geom_vline(xintercept = 0) +
  scale_fill_manual(values = DIRECTION_COLOURS) +
  scale_x_continuous(labels = abs) +
  labs(x = "significant proteins (down | up)", y = NULL, fill = NULL)
```

## Volcano plots {.tabset .tabset-dropdown}
Pick a contrast. Dashed lines mark the significance cutoffs; the most significant proteins are labelled.

```{r, results = "asis", fig.height = 5.5}
for (name in names(de)) {
  cat("\n\n### ", name, "\n\n", sep = "")
  print(volcano_plot(de[[name]], criterion) + labs(subtitle = str_glue("up = higher in {pairs[[name]][1]}")))
}
```

## Top hits
Intensities of the `r length(top_ids)` most significant proteins, z-scored per protein (red = above the protein's mean across samples). Samples are grouped by condition.
`r if (length(top_ids) < 2) "Not shown: fewer than two significant proteins."`

```{r, fig.height = heatmap_height, eval = length(top_ids) >= 2}
top <- proteins |> filter(protein_id %in% top_ids)
heatmap <- as.matrix(top[samples$sample])
rownames(heatmap) <- make.unique(top$gene)

pheatmap::pheatmap(
  heatmap,
  scale = "row",
  color = colorRampPalette(c("#2166AC", "white", "#B2182B"))(100),
  cluster_cols = FALSE,
  gaps_col = head(cumsum(rle(samples$condition)$lengths), -1),
  annotation_col = data.frame(condition = conditions, row.names = samples$sample),
  annotation_colors = list(condition = condition_colours),
  annotation_names_col = FALSE,
  show_colnames = FALSE,
  border_color = NA,
  fontsize = 7
)
```

## p-value distributions
A healthy contrast shows a flat histogram with a peak near 0 (the real differences). A hump in the middle or near 1 points to a problem with the model, such as an unaccounted batch or pairing.

```{r, fig.height = pvalue_height}
de |>
  map(\(x) select(x, pvalue)) |>
  list_rbind(names_to = "contrast") |>
  mutate(contrast = fct_inorder(contrast)) |>
  ggplot(aes(pvalue)) +
  geom_histogram(breaks = seq(0, 1, 0.05), fill = "grey50") +
  facet_wrap(vars(contrast), ncol = 4, scales = "free_y") +
  theme(strip.text = element_text(size = 7)) +
  labs(x = "p-value", y = "proteins")
```

## How to read this

- **A_vs_B**: positive log2FC means higher in A. A is the condition listed first in `samples.tsv`; to flip a comparison, reorder the rows of `Input/<dataset>/samples.tsv`.
- **Significant**: `r describe_criterion(criterion)`. padj is the Benjamini-Hochberg adjusted p-value within each contrast.
- **Statistics**: one limma model for all conditions, so every contrast borrows variance information from all samples.
- **Paired samples**: when `samples.tsv` has a `block` column, samples with the same value (e.g. the same mouse) are treated as related, which removes subject-to-subject differences from the comparisons.
- **Tables**: all proteins with statistics in `de/all_proteins/`, significant ones in `de/<criterion>/significant/`.
