---
title: "Enrichment report"
output:
  html_document:
    toc: true
    toc_depth: 3
    mathjax: null
params:
  rmd: ""
---

```{r setup, include = FALSE}
# Gene sets + GSEA and ORA results of every flagged collection (scripts/gsea.R, scripts/ora.R) -> one enrichment report
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
source("scripts/common.R")

proteins <- read_tsv(snakemake@input$intensities, show_col_types = FALSE)
species  <- detect_species(proteins)
msigdb   <- snakemake@params$msigdb  # flagged collections, in the order of enrichment.tsv
criteria <- snakemake@params$criteria
gsea     <- unlist(snakemake@input$gsea)  # enrichment/<collection>/gsea, for collections flagged gsea
ora      <- unlist(snakemake@input$ora)   # enrichment/<collection>/ora/<criterion>, for collections flagged ora
folders  <- dirname(unlist(snakemake@input$gene_sets)) |> set_names(basename)
detected <- n_distinct(first_gene(proteins$gene))

collection_row <- function(collection) {
  folder    <- folders[[collection]]
  sets      <- read_tsv(file.path(folder, "gene_sets.tsv"), show_col_types = FALSE)
  annotated <- n_distinct(sets$gene)
  tibble(
    collection,
    source  = msigdb_source(msigdb[[collection]], species)$note,
    tested  = sets |> count(term) |> filter(between(n, MIN_SET_SIZE, MAX_SET_SIZE)) |> nrow(),
    genes   = str_glue("{annotated} of {detected} ({round(100 * annotated / detected)}%)"),
    methods = str_flatten_comma(c(if (file.path(folder, "gsea") %in% gsea) "GSEA", if (any(dirname(dirname(ora)) == folder)) "ORA"))
  )
}

overview <- tribble(
  ~item, ~value,
  "Dataset", snakemake@wildcards$dataset,
  "Species", species$name,
  "Sets tested if", str_glue("{MIN_SET_SIZE}-{MAX_SET_SIZE} of their genes detected"),
  "Enriched if", str_glue("padj < {ENRICH_PADJ}")
)

# One method of one collection: enriched sets per contrast, the overview when there is one, then one tab per contrast
section <- function(title, folder, intro, columns) {
  summary <- read_tsv(file.path(folder, "summary.tsv"), show_col_types = FALSE)
  cat("\n\n### ", title, " {.tabset .tabset-dropdown}\n\n", intro, "\n\n#### Enriched sets per contrast\n\n", sep = "")
  cat(knitr::kable(select(summary, all_of(columns)), col.names = names(columns)), sep = "\n")
  if (dir.exists(file.path(folder, "figures", "overview"))) {
    cat("\n\n#### All contrasts\n\n", figure_image(folder, "figures", "overview", ext = "svg"), "\n", sep = "")
  }
  for (i in seq_len(nrow(summary))) {
    cat("\n\n#### ", summary$label[i], "\n\n", figure_image(folder, "figures", summary$contrast[i], ext = "svg"), "\n", sep = "")
  }
}
```

## Summary

```{r}
knitr::kable(overview, col.names = NULL)
```

```{r}
names(msigdb) |>
  map(collection_row) |>
  list_rbind() |>
  knitr::kable(col.names = c("collection", "gene sets", "sets tested", "detected genes in any set", "methods"))
```

```{r, results = "asis"}
for (collection in names(msigdb)) {
  folder <- folders[[collection]]
  cat("\n\n## ", collection, "\n\n", msigdb_source(msigdb[[collection]], species)$note, "\n\n", sep = "")
  if (file.path(folder, "gsea") %in% gsea) {
    section(
      "GSEA", file.path(folder, "gsea"),
      "All detected proteins ranked by the contrast's limma t statistic (and by |t| for mixed sets); no significance cutoff. Pick a contrast from the menu.",
      c(contrast = "label", "file name" = "contrast", "sets up" = "up", "sets down" = "down", "sets mixed" = "mixed",
        "sets tested" = "tested")
    )
  }
  for (dir in ora[dirname(dirname(ora)) == folder]) {
    criterion <- basename(dir)
    section(
      str_glue("ORA: {criterion}"), dir,
      str_glue("Significant proteins ({describe_criterion(criteria[[criterion]])}): up, down, and both together for mixed sets. Pick a contrast from the menu."),
      c(contrast = "label", "file name" = "contrast", "proteins up" = "genes_up", "sets up" = "up",
        "proteins down" = "genes_down", "sets down" = "down", "sets mixed" = "mixed")
    )
  }
}
```

## How to read this

- **GSEA** (gene set enrichment analysis, fgsea): all detected proteins are ranked by the limma moderated t of the contrast A vs B. A set with a positive NES (normalized enrichment score) has its proteins near the top, i.e. higher in A; negative NES, higher in B. It needs no significance cutoff and picks up sets whose proteins shift together, even when few of them are significant on their own.
- **ORA** (over-representation analysis): are the significant up (or down, or all) proteins found in a set more often than expected from all detected proteins in the collection? Hypergeometric test, one per significance criterion. **Fold enrichment** = share of the significant proteins that fall in the set / share of all detected proteins that do. Proteins up and down only count when they belong to at least one set of the collection.
- **Mixed**: a set whose proteins go both ways (some up, some down) cancels out in GSEA and is diluted in ORA's up and down lists. So each method also tests both directions together: GSEA on |t|, ORA on all significant proteins. A set is **mixed** when it is enriched only in that combined test, not up or down alone, and its proteins (GSEA: the leading edge; ORA: the significant ones in the set) include both directions. The bars show the split.
- **padj**: Benjamini-Hochberg within each contrast (and direction, for ORA). A set is called enriched at padj < `r ENRICH_PADJ`. The figures show enriched sets only; every tested set is in the tables.
- **Genes**: a protein group counts as its first gene; sets are tested on the `r MIN_SET_SIZE`-`r MAX_SET_SIZE` of their genes detected in this dataset, so set sizes are smaller than in MSigDB.
- **Tables** in `enrichment/<collection>/`: `gene_sets.tsv` (the sets used), `gsea/A_vs_B.tsv` (every set: signed NES, padj, leading_edge = the genes driving the enrichment; any_* = the |t| run; call = up, down or mixed), `ora/<criterion>/A_vs_B.tsv` (sets sharing at least one significant protein, for the up, down and any = all significant proteins lists; genes = those proteins, n_up and n_down = how many go each way).
- **Figures**: each plot is also in its own folder with its data and a script to restyle it: `gsea/figures/<A_vs_B>/`, `gsea/figures/overview/`, `ora/<criterion>/figures/<A_vs_B>/`.
