---
title: "Proteomics QC report"
output:
  html_document:
    toc: true
    mathjax: null
params:
  rmd: ""
---

```{r setup, include = FALSE}
# QC tables and figures (scripts/qc.R) -> one QC report
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
options(knitr.kable.NA = "")
source("scripts/common.R")

qc         <- snakemake@input$qc
figures    <- file.path(qc, "figures")
overview   <- read_tsv(file.path(qc, "overview.tsv"), show_col_types = FALSE)
samples    <- read_tsv(file.path(qc, "sample_metrics.tsv"), show_col_types = FALSE)
conditions <- read_tsv(file.path(qc, "condition_metrics.tsv"), show_col_types = FALSE)

complete <- overview$value[overview$item == "Proteins in every sample"]
has_figure <- \(name) dir.exists(file.path(figures, name))
```

## Summary

```{r}
knitr::kable(overview, col.names = NULL)
```

`r if (all(samples$missing_pct == 0)) "No missing values: the table was most likely imputed by the facility, so completeness says nothing about data quality here. Look at low_values_pct instead: imputed values usually land in the extreme low tail."`

## Samples

```{r}
knitr::kable(samples, digits = 3)
```

## Conditions

```{r}
knitr::kable(conditions, digits = 3)
```

## Plots
Each plot is also in `qc/figures/<plot>/` next to this report, with its data and a script to restyle it.

### Intensity distributions
Boxes should sit at similar heights; a shifted box means a loading or normalization problem.

`r figure_image(figures, "intensities")`

### Contaminants
Dashed line = dataset median.

`r figure_image(figures, "contaminants")`

### Sample correlation
Samples are ordered by similarity (hierarchical clustering on 1 − Pearson r), and the tree shows which samples group together. The colour bar marks each sample's condition: replicates of one condition should end up next to each other.

`r figure_image(figures, "correlation")`

### PCA
Replicates should cluster by condition. Flagged samples are labelled with their replicate number. Based on `r complete` proteins measured in every sample.

`r if (has_figure("pca")) figure_image(figures, "pca")`

### PCA per pair of conditions
Each panel is a separate PCA of just two conditions, so a group that dominates the PCA above cannot hide their differences. Separation along PC1 means the two conditions differ more than their replicates do. Points are labelled with replicate numbers.
`r if (!has_figure("pca_pairs")) "Not shown: with two conditions this is the same as the PCA above."`

`r if (has_figure("pca_pairs")) figure_image(figures, "pca_pairs")`

### Reproducibility within conditions
Coefficient of variation of each protein across the replicates of a condition (lower = more reproducible).

`r figure_image(figures, "cv")`

## How to read this

- **few proteins**: far fewer proteins quantified than the other samples (failed or low-input run).
- **high contaminants**: an unusually large share of intensity from cRAP proteins such as keratins, trypsin or serum albumin (handling or sample prep).
- **intensity shift**: median intensity more than `r MAX_MEDIAN_SHIFT` log2 (2-fold) away from the typical sample (loading or normalization problem).
- **low replicate correlation**: agrees less with its own replicates than samples usually do in this dataset (possible outlier or sample swap; check the PCA).

Not flagged, for context: **low_values_pct** is the share of a sample's values in the dataset's extreme low tail (mostly imputed, i.e. undetected, proteins), and **cor_all** is the sample's median correlation with every other sample.

"Unusual" means more than `r OUTLIER_MADS` median absolute deviations worse than the dataset median. A flag is a prompt to look, not a verdict.
To drop a sample, write a reason in the `exclude` column of `Input/<dataset>/samples.tsv` and run the pipeline again.
