---
title: "Proteomics QC report"
output:
  html_document:
    toc: true
    mathjax: null
params:
  rmd: ""
---

```{r setup, include = FALSE}
# Clean intensities + sample sheet (+ raw table for contaminants) -> one QC report
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 9)
source("scripts/common.R")

# A sample is flagged when it stands out from the rest of the dataset
OUTLIER_MADS     <- 3  # proteins, contaminants, replicate correlation: > 3 MADs worse than the median
MAX_MEDIAN_SHIFT <- 1  # median intensity: > 1 log2 (2-fold) away from the typical sample

# side = +1: high values are bad; side = -1: low values are bad
is_outlier <- function(x, side) side * (x - median(x, na.rm = TRUE)) > OUTLIER_MADS * mad(x, na.rm = TRUE)

samples  <- read_tsv(snakemake@input$samples, show_col_types = FALSE) |>
  mutate(condition = fct_inorder(condition))  # keep sample-sheet order in tables and plots
proteins <- read_tsv(snakemake@input$intensities, show_col_types = FALSE)
raw      <- read_table(snakemake@input$table)
sheet    <- read_tsv(snakemake@input$sheet, col_types = cols(.default = "c"))

log2i <- as.matrix(proteins[samples$sample])  # log2 intensities, proteins x samples

# Contaminants: share of each sample's total intensity that comes from cRAP proteins
crap <- str_starts(raw$protein_id, "cRAP-")
raw_linear <- 2^as.matrix(raw[samples$column])
contaminants <- colSums(raw_linear[crap, , drop = FALSE], na.rm = TRUE) / colSums(raw_linear, na.rm = TRUE)

# Sample-to-sample correlation; `others` blanks the diagonal so a sample is not compared to itself
cors <- cor(log2i, use = "pairwise.complete.obs")
others <- cors
diag(others) <- NA
replicate_cors <- ifelse(outer(samples$condition, samples$condition, "=="), others, NA)

# Extreme low values (below the dataset's lower boxplot whisker) usually mark imputed, i.e. undetected, proteins
quartiles <- quantile(log2i, c(0.25, 0.75), na.rm = TRUE)
low_cutoff <- quartiles[[1]] - 1.5 * diff(quartiles)

qc <- samples |>
  select(sample, condition, replicate) |>
  mutate(
    quantified       = colSums(!is.na(log2i)),
    missing_pct      = 100 * colMeans(is.na(log2i)),
    low_values_pct   = 100 * colMeans(log2i < low_cutoff, na.rm = TRUE),
    contaminants_pct = 100 * contaminants,
    median_shift     = apply(log2i - apply(log2i, 1, median, na.rm = TRUE), 2, median, na.rm = TRUE),
    cor_within       = apply(replicate_cors, 2, median, na.rm = TRUE),
    cor_all          = apply(others, 2, median, na.rm = TRUE),
    flags = str_c(
      if_else(is_outlier(quantified, -1), "few proteins; ", "", missing = ""),
      if_else(is_outlier(contaminants_pct, +1), "high contaminants; ", "", missing = ""),
      if_else(abs(median_shift) > MAX_MEDIAN_SHIFT, "intensity shift; ", "", missing = ""),
      if_else(is_outlier(cor_within, -1), "low replicate correlation; ", "", missing = "")
    ) |> str_remove("; $")
  )

# Long table (one row per protein x sample) for per-condition CVs and plots
long <- proteins |>
  select(protein_id, all_of(samples$sample)) |>
  pivot_longer(-protein_id, names_to = "sample", values_to = "log2") |>
  left_join(select(samples, sample, condition), by = "sample") |>
  mutate(sample = fct_inorder(sample))

cv <- long |>
  summarise(cv = 100 * sd(2^log2, na.rm = TRUE) / mean(2^log2, na.rm = TRUE), .by = c(condition, protein_id))

conditions <- qc |>
  summarise(replicates = n(), median_cor_within = median(cor_within), .by = condition) |>
  left_join(summarise(cv, median_cv_pct = median(cv, na.rm = TRUE), .by = condition), by = "condition")

# Dataset summary
contaminant_names <- str_remove(if ("protein_name" %in% names(raw)) raw$protein_name else raw$protein_id, "^cRAP-")
top_contaminants <- contaminant_names[crap][order(-rowSums(raw_linear[crap, , drop = FALSE], na.rm = TRUE))]
excluded <- sheet |> filter(if_any(any_of("exclude"), \(x) !is.na(x)))

overview <- tribble(
  ~item, ~value,
  "Dataset", snakemake@wildcards$dataset,
  "Samples", str_glue("{nrow(samples)} in {n_distinct(samples$condition)} conditions"),
  "Proteins", as.character(nrow(proteins)),
  "Missing values", str_glue("{round(100 * mean(is.na(log2i)), 1)}%"),
  "Single-peptide proteins", if ("n_sequences" %in% names(proteins)) {
    str_glue("{round(100 * mean(proteins$n_sequences == 1, na.rm = TRUE), 1)}%")
  } else "not available",
  "Ambiguous protein groups", as.character(sum(str_detect(proteins$protein_id, ";"))),
  "Contaminants removed", str_glue("{sum(crap)} cRAP proteins (top: {str_flatten_comma(head(top_contaminants, 5))})"),
  "Excluded samples", if (nrow(excluded)) str_flatten_comma(str_glue("{excluded$column} ({excluded$exclude})")) else "none",
  "Flagged samples", str_glue("{sum(qc$flags != '')} of {nrow(qc)}")
)

# PCA on the proteins measured in every sample of x: PC1/PC2 per sample + % variance of each
pca_scores <- function(x) {
  pca <- prcomp(t(x[complete.cases(x), , drop = FALSE]))
  variance <- round(100 * pca$sdev^2 / sum(pca$sdev^2), 1)
  tibble(sample = colnames(x), PC1 = pca$x[, 1], PC2 = pca$x[, 2], var1 = variance[1], var2 = variance[2])
}
complete <- log2i[complete.cases(log2i), , drop = FALSE]
n_pairs <- choose(nlevels(samples$condition), 2)

# Same colour per condition in every plot (ggplot's default palette)
condition_colours <- set_names(scales::hue_pal()(nlevels(samples$condition)), levels(samples$condition))

# Plot heights grow with the number of samples / conditions / condition pairs
sample_height    <- max(4, 0.18 * nrow(samples))
condition_height <- max(2.5, 0.45 * nrow(conditions))
pair_height      <- 3 * ceiling(n_pairs / 3)
```

## Summary

```{r}
knitr::kable(overview, col.names = NULL)
```

`r if (!anyNA(log2i)) "No missing values: the table was most likely imputed by the facility, so completeness says nothing about data quality here. Look at low_values_pct instead: imputed values usually land in the extreme low tail."`

## Samples

```{r}
knitr::kable(qc, digits = 3)
```

## Conditions

```{r}
knitr::kable(conditions, digits = 3)
```

## Plots

### Intensity distributions
Boxes should sit at similar heights; a shifted box means a loading or normalization problem.

```{r, fig.height = sample_height}
ggplot(long, aes(log2, fct_rev(sample), fill = condition)) +
  geom_boxplot(outlier.size = 0.3, na.rm = TRUE) +
  labs(x = "log2 intensity", y = NULL)
```

### Contaminants
Dashed line = dataset median.

```{r, fig.height = sample_height}
ggplot(qc, aes(contaminants_pct, fct_rev(fct_inorder(sample)), fill = condition)) +
  geom_col() +
  geom_vline(xintercept = median(qc$contaminants_pct), linetype = "dashed") +
  labs(x = "% of total intensity from cRAP contaminants", y = NULL)
```

### Sample correlation
Samples are ordered by similarity (hierarchical clustering on 1 − Pearson r), and the tree shows which samples group together. The colour bar marks each sample's condition: replicates of one condition should end up next to each other.

```{r, fig.height = sample_height + 1.5}
annotation <- data.frame(condition = samples$condition, row.names = samples$sample)
distance <- as.dist(1 - cors)

pheatmap::pheatmap(
  cors,
  color = hcl.colors(100, "viridis"),
  clustering_distance_rows = distance,
  clustering_distance_cols = distance,
  clustering_method = "average",
  annotation_row = annotation,
  annotation_col = annotation,
  annotation_colors = list(condition = condition_colours),
  annotation_names_row = FALSE,
  annotation_names_col = FALSE,
  border_color = NA,
  fontsize = 7
)
```

### PCA
Replicates should cluster by condition. Flagged samples are labelled with their replicate number. Based on `r nrow(complete)` proteins measured in every sample.

```{r, fig.height = 6, eval = nrow(complete) >= 2}
pca <- pca_scores(log2i)

qc |>
  left_join(pca, by = "sample") |>
  ggplot(aes(PC1, PC2, colour = condition)) +
  geom_point(size = 2.5) +
  geom_text(aes(label = if_else(flags == "", "", str_c("R", replicate))), size = 3, vjust = -0.9, show.legend = FALSE) +
  scale_colour_manual(values = condition_colours) +
  scale_x_continuous(expand = expansion(mult = 0.1)) +
  labs(x = str_glue("PC1 ({pca$var1[1]}%)"), y = str_glue("PC2 ({pca$var2[1]}%)"))
```

### PCA per pair of conditions
Each panel is a separate PCA of just two conditions, so a group that dominates the PCA above cannot hide their differences. Separation along PC1 means the two conditions differ more than their replicates do. Points are labelled with replicate numbers.
`r if (n_pairs < 2) "Not shown: with two conditions this is the same as the PCA above."`

```{r, fig.height = pair_height, eval = n_pairs >= 2}
combn(levels(samples$condition), 2, simplify = FALSE) |>
  map(\(pair) {
    pca_scores(log2i[, samples$condition %in% pair, drop = FALSE]) |>
      mutate(pair = str_glue("{pair[1]} vs {pair[2]}\nPC1 {var1}%, PC2 {var2}%"))
  }) |>
  list_rbind() |>
  left_join(select(samples, sample, condition, replicate), by = "sample") |>
  mutate(pair = fct_inorder(pair)) |>
  ggplot(aes(PC1, PC2, colour = condition)) +
  geom_point(size = 1.8) +
  geom_text(aes(label = str_c("R", replicate)), size = 2.5, vjust = -0.8, show.legend = FALSE) +
  facet_wrap(vars(pair), scales = "free", ncol = 3) +
  scale_colour_manual(values = condition_colours) +
  scale_x_continuous(expand = expansion(mult = 0.15)) +
  scale_y_continuous(expand = expansion(mult = 0.15)) +
  theme(legend.position = "top", axis.text = element_text(size = 6))
```

### Reproducibility within conditions
Coefficient of variation of each protein across the replicates of a condition (lower = more reproducible).

```{r, fig.height = condition_height}
ggplot(cv, aes(cv, fct_rev(condition), fill = condition)) +
  geom_boxplot(outlier.shape = NA, na.rm = TRUE, show.legend = FALSE) +
  coord_cartesian(xlim = c(0, 100)) +
  labs(x = "CV % per protein", y = NULL)
```

## How to read this

- **few proteins**: far fewer proteins quantified than the other samples (failed or low-input run).
- **high contaminants**: an unusually large share of intensity from cRAP proteins such as keratins, trypsin or serum albumin (handling or sample prep).
- **intensity shift**: median intensity more than `r MAX_MEDIAN_SHIFT` log2 (2-fold) away from the typical sample (loading or normalization problem).
- **low replicate correlation**: agrees less with its own replicates than samples usually do in this dataset (possible outlier or sample swap; check the PCA).

Not flagged, for context: **low_values_pct** is the share of a sample's values in the dataset's extreme low tail (mostly imputed, i.e. undetected, proteins), and **cor_all** is the sample's median correlation with every other sample.

"Unusual" means more than `r OUTLIER_MADS` median absolute deviations worse than the dataset median. A flag is a prompt to look, not a verdict.
To drop a sample, write a reason in the `exclude` column of `Input/<dataset>/samples.tsv` and run the pipeline again.
